---
title: "Red Run Study Sites"
author: "Frank"
date: "`r format(Sys.Date())`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=10, fig.height=8)
```

```{css, echo=FALSE}
h1, h4 {
  text-align: center;
  color: black;
  font-weight: bold;
}
```

# Introduction

These are the codes I used to create an interactive map of my study area/sites in Red Run Watershed. All credit goes to Nicole Williamson - a graduate MS student who shared with me her codes so that I could replicate them. Check out her work [here.](https://williamson-nicole.github.io/Towson_Williamson/ "Nicole the GOAT")

## Install and load the package "leaflet"

```{r leaflet, message=F,warning=F}
library (leaflet)
```

## Load latitude and longitude data

```{r load data,warning=F,message=F}
library(readxl)
RR_benthic <- read_excel("F:/Thesis stuff/Data/Land cover data/Biological Sites/Benthos Site.xls")
```

## Load RR watershed's and subwatersheds' boundary polygon files

```{r map creation, message=F,warning=F,results='hide'}
library(rgdal) #load the package
rr_boundary <- readOGR("F:/Thesis stuff/Data/Land cover data/RR Benthic Sites Raster/RR Site 1/layers/globalwatershed.shp",GDAL1_integer64_policy = T)
b7_boundary <- readOGR("F:/Thesis stuff/Data/Land cover data/RR Benthic Sites Raster/RR Site 7/layers/globalwatershed.shp")
b9a_boundary <- readOGR("F:/Thesis stuff/Data/Land cover data/RR Benthic Sites Raster/RR Site 9A/layers/globalwatershed.shp")
b9_boundary <- readOGR("F:/Thesis stuff/Data/Land cover data/RR Benthic Sites Raster/RR Site 9/layers/globalwatershed.shp")
```

## Create the map

```{r, message=F,warning=F}
leaflet(RR_benthic) %>% 
  addPolygons(data=rr_boundary,weight=5,col = '#b10026', smoothFactor = 0.5,
    opacity = 1.0, fillOpacity = 0.5) %>% 
  addPolygons(data=b7_boundary,weight=5,col = '#feb14c', smoothFactor = 0.5,
    opacity = 1.0, fillOpacity = 0.5) %>% 
  addPolygons(data=b9a_boundary,weight=5,col = '#fc4d2a', smoothFactor = 0.5,
    opacity = 1.0, fillOpacity = 0.5) %>% 
  addPolygons(data=b9_boundary,weight=5,col = '#ffeea1', smoothFactor = 0.5,
    opacity = 1.0, fillOpacity = 0.5) %>% 
  addProviderTiles(providers$Esri.NatGeoWorldMap) %>%
  addTiles() %>% addScaleBar() %>% addMarkers(lng=~Long,lat=~Lat,popup = ~ Site) 
```

