---
title: "Multiple Regression of WC and NMDS scores"
author: "Frank"
date: "`r format(Sys.Date())`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=10, fig.height=8)
```

```{css, echo=FALSE}
h1, h4 {
  text-align: center;
  color: black;
  font-weight: bold;
}

/* Whole document: */
body{
  font-family: Times New Roman;
  font-size: 16pt;
}
/* Headers */
h1,h2,h3,h4,h5,h6{
  font-size: 24pt;
}

<style type="text/css">
body, td {
   font-size: 16px;
}
code.r{
  font-size: 16px;
}
pre {
  font-size: 16px
}
</style>
  
```

# Introduction

These are the codes I used to run the multiple regression analysis on the correlations between axis 1/2 scores and water chemistry measurements from B1, B5, B7, B9, and B9A, which correspond to five water sampling stations designated by the Baltimore County.

## Edit and organize data

I am going to use data from data.scores.dplyr (Check my NMDS Combined Density) and from wc (Check my PCA Water Chemistry). I will use *dplyr* package to combine and filter data to fit purpose.

### Data for later use

```{r message=FALSE,warning=F}
library(readxl)
wc <- read_excel("F:/Thesis stuff/Data Analysis/Part a (87-09)/PCA (Cl, Ca, SC)/Data for PCA.xlsx")
data.scores.dplyr <- read_excel("C:/Users/tiena/OneDrive/Documents/thesis-codes/Part a (87-09)/Multiple regression between WC and axis scores/NMDS scores-dplyr.xlsx")
```


```{r message=F,warning=F}
library(dplyr)
wc_dplyr <- wc %>% filter(!row_number() %in% c(6,12)) %>% group_by(Year) %>% arrange(match(Site,c("RR1","RR2","RR5","RR4","RR3"))) 

#delete row 6 and 12 that contains data from RR6, which does not match with any benthic site 
#rearrange rows so that water quality sites will match benthic sites

mp_data <- data.scores.dplyr %>% filter(!row_number() %in% c(2,3, 4,7,10,12,13,14,17,20)) %>% arrange(match(Site,c("B1","B5","B7","B9","B9A"))) %>% relocate(Site, Year, Subwatershed) %>% mutate(TSS=wc_dplyr$TSS,TKN=wc_dplyr$TKN,TP=wc_dplyr$TP,
                           Cl=wc_dplyr$Cl,Ca=wc_dplyr$Ca,SC=wc_dplyr$SC) 
#remove rows that contain data from benthic sites that do not match with water quality sites 
#rearrange sites
#add columns of WC data using mutate function
#relocate Site, Year, Subwatershed columns to the front; you can use relocate(column a, .after=column b) => this will put column a after column b
```

Now that the data are ready to be used. The rest is easy and straighforward. 

## Conduct multiple regression

### NMDS Axis 1 score vs WC

```{r message=F,warning=F}
axis1_mp_salinity <- lm(data=mp_data,NMDS1~Cl+SC)
summary(axis1_mp_salinity)

axis1_mp_nutrient <- lm(data=mp_data,NMDS1~TKN+TP+TSS)
summary(axis1_mp_nutrient)

axis1_mp <- lm(data=mp_data, NMDS1~Cl+SC+Ca+TKN+TP+TSS)
summary(axis1_mp) #Ca is significant in this model
```

### NMDS Axis 2 score vs WC

```{r message=F,warning=F}
axis2_mp_salinity <- lm(data=mp_data,NMDS2~Cl+SC)
summary(axis2_mp_salinity)

axis2_mp_nutrient <- lm(data=mp_data,NMDS2~TKN+TP+TSS)
summary(axis2_mp_nutrient) #Intercept significant

axis2_mp <- lm(data=mp_data, NMDS2~Cl+SC+Ca+TKN+TP+TSS)
summary(axis2_mp)
```

# References

1. Jari Oksanen, Gavin L. Simpson, F. Guillaume Blanchet, Roeland Kindt, Pierre Legendre, Peter R. Minchin,
  R.B. O'Hara, Peter Solymos, M. Henry H. Stevens, Eduard Szoecs, Helene Wagner, Matt Barbour, Michael
  Bedward, Ben Bolker, Daniel Borcard, Gustavo Carvalho, Michael Chirico, Miquel De Caceres, Sebastien
  Durand, Heloisa Beatriz Antoniazi Evangelista, Rich FitzJohn, Michael Friendly, Brendan Furneaux,
  Geoffrey Hannigan, Mark O. Hill, Leo Lahti, Dan McGlinn, Marie-Helene Ouellette, Eduardo Ribeiro Cunha,
  Tyler Smith, Adrian Stier, Cajo J.F. Ter Braak and James Weedon (2022). vegan: Community Ecology Package.
  R package version 2.6-2. https://CRAN.R-project.org/package=vegan
2. Hadley Wickham, Romain François, Lionel Henry and Kirill Müller (2022). dplyr: A Grammar of Data
  Manipulation. R package version 1.0.8. https://CRAN.R-project.org/package=dplyr
3. Hadley Wickham. ggplot2: Elegant Graphics for Data Analysis. Springer-Verlag New York, 2016.