---
title: "PCA of Metric Data - Part B: 2009-2019"
author: "Frank"
date: "`r format(Sys.Date())`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=10, fig.height=8)
```

```{css, echo=FALSE}
h1, h4 {
  text-align: center;
  color: black;
  font-weight: bold;
}

/* Whole document: */
body{
  font-family: Times New Roman;
  font-size: 16pt;
}
/* Headers */
h1,h2,h3,h4,h5,h6{
  font-size: 24pt;
}

<style type="text/css">
body, td {
   font-size: 16px;
}
code.r{
  font-size: 16px;
}
pre {
  font-size: 16px
}
</style>
  
```

# Introduction

These are the codes I used to run PCA for my metric data calculated using the MBSStools package (Leppo, 2021). Again, I welcome any feedback if there is any to further improve my codes.

## Load data

Data can be found [here.](https://github.com/tquach123/thesis-codes "Frank's GitHub")

```{r message=F, warning=F}
library(readxl)
metric_data <- read_excel("~/thesis-codes/Part b (09-19)/Main Matrices No Zero.xlsx", 
    sheet = "MetricIBI")
```

## Make a matrix from raw data

```{r message=F,warning=F}
metric_data_su <- metric_data[,4:ncol(metric_data)] #exclude the first column that contains SU information
metric.pca <- princomp(metric_data_su, cor=T)
summary(metric.pca)
metric.pca$loadings #check the loading of each variable on each PC
```

## Plot Screeplot

This step is optional. The previous step already showed you how many percentage of variance is explained by each PC. This is to visualize by using package factoextra (Kassambara and Mundt, 2020).

```{r message=F,warning=F}
library(factoextra)
fviz_eig(metric.pca)
```

## Add columns of information to the result

Another way to add columns to the existing dataset is to use pipes from the dplyr package (Wickham et al., 2022).

```{r message=F, warning=F}
library(dplyr) #load the package
data.scores <- as.data.frame(metric.pca$scores[,1:2]) #extract NMDS scores
data.scores.dplyr <- data.scores %>% mutate(Site = metric_data$Site,Year=metric_data$Year,
                                            Subwatershed=metric_data$Subwatershed)
head(data.scores.dplyr)
```

## Plot PCA graph

```{r message=F, warning=F}
library(ggplot2)

ggplot(data.scores.dplyr,aes(Comp.1, Comp.2))+ 
    geom_point(size = 4,aes( shape = as.character(Year))) +
    theme(axis.text.y = element_text(colour = "black", size = 12, face = "bold"), 
    axis.text.x = element_text(colour = "black", face = "bold", size = 12), 
    legend.text = element_text(size = 12, face ="bold", colour ="black"), 
    legend.position = "right", axis.title.y = element_text(face = "bold", size = 14), 
    axis.title.x = element_text(face = "bold", size = 14, colour = "black"), 
    legend.title = element_text(size = 14, colour = "black", face = "bold"), 
    panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2),
    legend.key=element_blank(),
    plot.title = element_text(color = "black", size = 30, face = "bold", hjust = 0.5)) + 
    labs(
    title = "PCA graph of metrics and IBI scores") + 
    theme(axis.title.x = element_text(margin=margin(t=10)), #add margin to x-axis title
        axis.title.y = element_text(margin=margin(r=10)))+
    labs(x = "PC1 (66.5%)", y = "PC2 (14.5%)", shape = "Year") +
    geom_text(aes(label=Site),hjust=-0.15, vjust=1)
```




# References
1. Erik W. Leppo (2021). MBSStools: MBSS tool suite for calculations and data manipulation. R package
  version 1.1.0.9053. https://github.com/leppott/MBSStools
2. Alboukadel Kassambara and Fabian Mundt (2020). factoextra: Extract and Visualize the Results of
  Multivariate Data Analyses. R package version 1.0.7. https://CRAN.R-project.org/package=factoextra
3. Hadley Wickham, Romain François, Lionel Henry and Kirill Müller (2022). dplyr: A Grammar of Data
  Manipulation. R package version 1.0.8. https://CRAN.R-project.org/package=dplyr
