---
title: "Discharge Analysis (Chap 3)"
author: "Frank"
date: "`r format(Sys.Date())`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=15, fig.height=10)
```

```{css, echo=FALSE}
h1, h4 {
  text-align: center;
  color: black;
  font-weight: bold;
}

/* Whole document: */
body{
  font-family: Times New Roman;
  font-size: 16pt;
}
/* Headers */
h1,h2,h3,h4,h5,h6{
  font-size: 24pt;
}

<style type="text/css">
body, td {
   font-size: 16px;
}
code.r{
  font-size: 16px;
}
pre {
  font-size: 16px
}
</style>
  
```

# Introduction

To investigate the impacts of climate change on stream hydrology, it is important to examine the long-term trends. However, in urban watersheds, the question is then how to disentangle the effects of urbanization and climate change on hydrology. 

# Villa Nova Flow duration curve

I first want to know how flow duration curves differ during different periods of time at VN gauging station.

## Load libraries, set working directory, & load data

```{r message=F, warning=F}
## load necessary libraries
library(lubridate) # Manipulate dates

# USGS library
library(EGRET)

# Required for some of functions below that EGRET uses and to run the functions pasted in at the bottom
library(rkt)
library(zyp)
library(Kendall)
```

## Read data - gauging site number, Q data from NWIS

```{r message=F,warning=F}
# Set gauging site number
siteNumber <- '01589300' # Gwynns Falls at Villa Nova MD

# Set start & end dates for period
startDate <- '1957-02-28'
endDate <- '2019-12-31' # through end of the study period, 2019

# Read in daily discharge (Q) data from USGS National Water Info System (NWIS)
Daily <- readNWISDaily(siteNumber, '00060', startDate, endDate)
# 00060 is the USGS parameter code for stream discharge

# Read in information about site & set short name - note ** interactive
INFO <- readNWISInfo(siteNumber, '00060') 
# Set short name to 'Gwynns Falls-Villa Nova'
# Set station abbreviation to 'GWVN'
# Set 'Stream flow, mean. daily' to 'DailyQ_cfs'
# Set abbreviation for discharge units (ft3/s) to 'cfs'

# Create eList -- EGRET variable
eList <- as.egret(INFO, Daily, NA, NA)
```

## Pull out max discharge in a year and Plot duration curves

```{r message=F,warning=F}
library(dplyr)
library(ggplot2)

max_annualQ_early <- Daily %>% group_by(waterYear) %>% 
  slice(which.max(Q)) %>% select(waterYear,Q) %>%
  filter(waterYear <= 1988) %>% filter(waterYear != 1957) %>%
  ungroup() %>% mutate(specific_Q = Q/32.5, rank = rank(-Q), ri = (length(Q)+1)/rank,
                       percent_exceed = (1/ri)*100, time = "1958-1988") %>%
  as.data.frame() 

max_annualQ_late <- Daily %>% group_by(waterYear) %>% 
  slice(which.max(Q)) %>% select(waterYear,Q) %>%
  filter(waterYear >= 1996 ) %>% filter(waterYear != 2020) %>%
  ungroup() %>% mutate(specific_Q = Q/32.5, rank = rank(-Q), ri = (length(Q)+1)/rank,
                       percent_exceed = (1/ri)*100, time= "1997-2019") %>%
  as.data.frame() 

max_annualQ <- full_join(max_annualQ_early,max_annualQ_late) %>% 
   ggplot(aes(x=percent_exceed, y=Q, group=time,color=time)) +
    geom_line(size=4) +
    scale_color_manual(values=c("#1b9e77","#e7298a")) +
    scale_x_log10(limits=c(1,100),
                  breaks=c(1,10,30,50,70,90)) +
    scale_y_log10(limits=c(1,300),breaks=c(1,10,100)) +
    #coord_trans(x="log10", y= "log10") + 
    #scale_x_continuous(breaks=c(1,10,30,50,70,90)) +
    #scale_y_continuous(breaks=c(1,10,100)) +
    #scale_x_continuous(trans="log10") +
    #scale_y_continuous(trans="log10") +
    annotation_logticks(sides="trbl") +
    theme(axis.text.y = element_text(colour = "black", size = 20, face = "bold"), 
    axis.text.x = element_text(colour = "black", face = "bold", size = 20), 
    legend.text = element_text(size = 18, face ="bold", colour ="black"), 
    legend.position = "right", axis.title.y = element_text(face = "bold", size = 14), 
    axis.title.x = element_text(face = "bold", size = 25, colour = "black"), 
    axis.title.y.left = element_text(face = "bold", size = 25, colour = "black"),
    legend.title = element_text(size = 20, colour = "black", face = "bold"), 
    panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2),
    legend.key=element_blank(),
    plot.title = element_text(color = "black", size = 30, face = "bold", hjust = 0.5)) + 
    labs(
    title = "Percent exceedance of maximum annual Q at GWVN",
    x = "Percent exceedance (%)",color="Period") +
    ylab (bquote('Discharge '(ft^3/s))) +
    theme(axis.title.x = element_text(margin=margin(t=10)), #add margin to x-axis title
        axis.title.y = element_text(margin=margin(r=10)))

ggsave("Max Annual Q Percent Exceedance at GWVN.jpeg",max_annualQ,height=13, width=25)

#stream disturbance regime: Stream Ecology by Allan JD
# relate to trout? check for data on MBSS; tie back to fish and food web => benthic macroinvertebrate

#air vs water temperature? 

```

## Daily Discharge

```{r message=F, warning=F}
library(dplyr)
library(ggplot2)
library(scales)

dailyQ_early <- Daily %>% select(Date,Q,waterYear) %>%
  filter(waterYear <= 1988) %>% filter(waterYear != 1957) %>%
  mutate(specific_Q = Q/32.5, rank = rank(-Q), ri = (length(Q)+1)/rank,
                       percent_exceed = (1/ri)*100, time = "1958-1988") %>%
  as.data.frame() 

dailyQ_late <- Daily %>% select(Date,Q,waterYear) %>%
  filter(waterYear >= 1996 ) %>% filter(waterYear != 2020) %>%
  mutate(specific_Q = Q/32.5, rank = rank(-Q), ri = (length(Q)+1)/rank,
                       percent_exceed = (1/ri)*100, time= "1997-2019") %>%
  as.data.frame() 

dailyQ_joined <- full_join(dailyQ_early,dailyQ_late)

dailyQ_duration <- full_join(dailyQ_early,dailyQ_late) %>% 
   ggplot(aes(x=percent_exceed, y=Q, group=time,color=time)) +
    geom_line(size=4) +
    scale_color_manual(values=c("#1b9e77","#e7298a")) +
    #scale_x_log10(limits=c(0.01,100),breaks=c(0.01,0.1,1,10,30,50,70,90)) +
    #scale_y_log10(limits=c(0.01,100),breaks=c(0.01,0.1,1,10,100)) +
    #coord_trans(x="log10", y= "log10") +
    scale_x_continuous(trans="log10", breaks=c(0.01,0.1,1,10,50,90))+
    scale_y_continuous(trans="log10", breaks=c(0.01,0.1,1,10,100)) +
    annotation_logticks(sides="trbl") +
    theme(axis.text.y = element_text(colour = "black", size = 20, face = "bold"), 
    axis.text.x = element_text(colour = "black", face = "bold", size = 20), 
    legend.text = element_text(size = 18, face ="bold", colour ="black"), 
    legend.position = "right", axis.title.y = element_text(face = "bold", size = 14), 
    axis.title.x = element_text(face = "bold", size = 25, colour = "black"), 
    axis.title.y.left = element_text(face = "bold", size = 25, colour = "black"),
    legend.title = element_text(size = 20, colour = "black", face = "bold"), 
    panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2),
    legend.key=element_blank(),
    plot.title = element_text(color = "black", size = 30, face = "bold", hjust = 0.5)) + 
    labs(
    title = "Percent exceedance of daily Q at GWVN",
    x = "Percent exceedance (%)",color="Period") +
    ylab (bquote('Discharge '(ft^3/s))) +
    theme(axis.title.x = element_text(margin=margin(t=10)), #add margin to x-axis title
        axis.title.y = element_text(margin=margin(r=10)))

ggsave("Daily Q Percent Exceedance at GWVN.jpeg",dailyQ_duration,height=13, width=25)
```

*Conclusion:* Results are consistent with [Colosimo and Wilcock (2002)](https://doi.org/10.1111/j.1752-1688.2007.00039.x "Colosimo and Wilcock 2002"). Earlier period (1958-1988) was shown to consistently have lower maximum annual Q than later period (1997-2019). This could be due to either climate change or urban development. The next section will attempt to disentangle the impacts of them.

# Relate Red Run to GWVN

Since USGS gauging station at Red Run near Owings Mills got installed recently in 2019, I will attempt to plot the specific discharge of the two stations and see how similar they are. The purpose is to infer about Red Run hydrology, while using GWVN as an estimate.

```{r message=F,warning=F}
#GWVN 

# Set gauging site number
siteNumber_GWVN <- '01589300' # Gwynns Falls at Villa Nova MD

# Set start & end dates for period
startDate_GWVN <- '2019-10-01'
endDate_GWVN <- '2022-09-30' # through end of the study period, 2019

# Read in daily discharge (Q) data from USGS National Water Info System (NWIS)
Daily_GWVN <- readNWISDaily(siteNumber_GWVN, '00060', startDate_GWVN, endDate_GWVN)
# 00060 is the USGS parameter code for stream discharge

# Read in information about site & set short name - note ** interactive
INFO_GWVN <- readNWISInfo(siteNumber_GWVN, '00060') 
# Set short name to 'Gwynns Falls-Villa Nova'
# Set station abbreviation to 'GWVN'
# Set 'Stream flow, mean. daily' to 'DailyQ_cfs'
# Set abbreviation for discharge units (ft3/s) to 'cfs'

# Create eList -- EGRET variable
eList_GWVN <- as.egret(INFO_GWVN, Daily_GWVN, NA, NA)
```

Red Run near Owings Mills (RROM)

Available daily Q: 2019-09-12 to present
Site code: 01589230 

```{r message=F,warning=F}
#GWVN 

# Set gauging site number
siteNumber_RROM <- '01589230' # Red Run near Owings Mills

# Set start & end dates for period
startDate_RROM <- '2019-10-01'
endDate_RROM <- '2022-09-30' # through end of the 2022 water year

# Read in daily discharge (Q) data from USGS National Water Info System (NWIS)
Daily_RROM <- readNWISDaily(siteNumber_RROM, '00060', startDate_RROM, endDate_RROM)
# 00060 is the USGS parameter code for stream discharge

# Read in information about site & set short name - note ** interactive
INFO_RROM <- readNWISInfo(siteNumber_RROM, '00060') 
# Set short name to 'Red Run-Owings Mills'
# Set station abbreviation to 'RROM'
# Set 'Stream flow, mean. daily' to 'DailyQ_cfs'
# Set abbreviation for discharge units (ft3/s) to 'cfs'

# Create eList -- EGRET variable
eList_RROM <- as.egret(INFO_RROM, Daily_RROM, NA, NA)
```


# Plot duration curves of the two watersheds

```{r message=F,warning=F}
library(dplyr)
library(ggplot2)

#GWVN 
                  
dailyQ_GWVN <- Daily_GWVN %>% select(MonthSeq,Q) %>%
  mutate(specific_Q = Q/32.5, rank = rank(-Q), ri = (length(Q)+1)/rank,
                       percent_exceed = (1/ri)*100, watershed = "GWVN") %>%
  as.data.frame() 

#RROM
dailyQ_RROM <- Daily_RROM %>%  select(MonthSeq,Q) %>%
  mutate(specific_Q = Q/19.1, rank = rank(-Q), ri = (length(Q)+1)/rank,
                       percent_exceed = (1/ri)*100, watershed= "RROM") %>%
  as.data.frame() 

dailyQ_RROM_GWVN <- full_join(dailyQ_GWVN,dailyQ_RROM) %>% 
   ggplot(aes(x=percent_exceed, y=specific_Q, group=watershed,color=watershed)) +
    geom_line(size=4) +
    scale_color_manual(values=c("#1b9e77","#e7298a")) +
    scale_x_continuous(trans="log10", breaks=c(0.01,0.1,1,10,50,100))+
    scale_y_continuous(trans="log10", breaks=c(0.001,0.01,0.1,1,10,100)) +
    #coord_trans(x="log10", y= "log10") + 
    #scale_x_continuous(breaks=c(1,10,30,50,70,90)) +
    #scale_y_continuous(breaks=c(1,10,100)) +
    #scale_x_continuous(trans="log10") +
    #scale_y_continuous(trans="log10") +
    annotation_logticks(sides="trbl") +
    theme(axis.text.y = element_text(colour = "black", size = 20, face = "bold"), 
    axis.text.x = element_text(colour = "black", face = "bold", size = 20), 
    legend.text = element_text(size = 18, face ="bold", colour ="black"), 
    legend.position = "right", axis.title.y = element_text(face = "bold", size = 14), 
    axis.title.x = element_text(face = "bold", size = 25, colour = "black"), 
    axis.title.y.left = element_text(face = "bold", size = 25, colour = "black"),
    legend.title = element_text(size = 20, colour = "black", face = "bold"), 
    panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2),
    legend.key=element_blank(),
    plot.title = element_text(color = "black", size = 30, face = "bold", hjust = 0.5)) + 
    labs(
    title = "Percent exceedance of mean daily specific Q at RROM and GWVN",
    x = "Percent exceedance (%)",color="Watershed") +
    ylab (bquote('Specific discharge '(ft^3/s~km^-2))) +
    theme(axis.title.x = element_text(margin=margin(t=10)), #add margin to x-axis title
        axis.title.y = element_text(margin=margin(r=10)))

ggsave("Mean Daily Specific Q Percent Exceedance at GWVN and RROM.jpeg",dailyQ_RROM_GWVN,height=13, width=25)
```

*Conclusion:* With specific discharge values of RROM and GWVN plotting next to each other, it was shown that GWVN should be a good representative of Red Run, since duration curves of the two watersheds are very similar to each other. 

# Flow duration curves of multiple watersheds

For this section, I am going to investigate the long-term trends of four watersheds, including Gwynns Falls at Villa Nova, Western Run, Gunpowder Falls at Hoffmanville, Piney Run at Dover, Morgan Run near Louisville, and Little Falls at Blue Mount.

```{r message=F,warning=F}
### 99. Functions from EGRET vignette page for Flow Trends ####
# http://usgs-r.github.io/EGRET/articles/streamflow_trend.html



########## this is the function you will use to make a single trend graph  ##############


plotFlowTrend <- function (eList, istat, startDate = NA, endDate = NA, 
                           paStart = 4, paLong = 12, window = 30, qMax = NA, 
                           printTitle = TRUE, tinyPlot = FALSE, 
                           customPar = FALSE, runoff = FALSE,
                           qUnit = 2, printStaName = TRUE, printPA = TRUE,
                           printIstat = TRUE, cex = 0.8, cex.axis = 1.1,
                           cex.main = 1.1, lwd = 2, col = "black", ...){
  localDaily <- getDaily(eList)
  localINFO <- getInfo(eList)
  localINFO$paStart <- paStart
  localINFO$paLong <- paLong
  localINFO$window <- window
  start <- as.Date(startDate)
  end <- as.Date(endDate)
  
  if(is.na(startDate)){
    start <- as.Date(localDaily$Date[1]) 
  } 
  
  if(is.na(endDate)){
    end <- as.Date(localDaily$Date[length(localDaily$Date)])
  }
  
  localDaily <- subset(localDaily, Date >= start & Date <= end)
  eList <- as.egret(localINFO,localDaily)
  localAnnualSeries <- makeAnnualSeries(eList)
  qActual <- localAnnualSeries[2, istat, ]
  qSmooth <- localAnnualSeries[3, istat, ]
  years <- localAnnualSeries[1, istat, ]
  Q <- qActual
  time <- years
  LogQ <- log(Q)
  mktFrame <- data.frame(time,LogQ)
  mktFrame <- na.omit(mktFrame)
  mktOut <- rkt::rkt(mktFrame$time,mktFrame$LogQ)
  zypOut <- zyp::zyp.zhang(mktFrame$LogQ,mktFrame$time)
  slope <- mktOut$B
  slopePct <- 100 * (exp(slope)) - 100
  slopePct <- format(slopePct,digits=2)
  pValue <- zypOut[6]
  pValue <- format(pValue,digits = 3)
  
  if (is.numeric(qUnit)) {
    qUnit <- qConst[shortCode = qUnit][[1]]
  } else if (is.character(qUnit)) {
    qUnit <- qConst[qUnit][[1]]
  }
  
  qFactor <- qUnit@qUnitFactor
  yLab <- qUnit@qUnitTiny
  
  if (runoff) {
    qActual <- qActual * 86.4/localINFO$drainSqKm
    qSmooth <- qSmooth * 86.4/localINFO$drainSqKm
    yLab <- "Runoff in mm/day"
  } else {
    qActual <- qActual * qFactor
    qSmooth <- qSmooth * qFactor
  }
  
  localSeries <- data.frame(years, qActual, qSmooth)
  
  
  yInfo <- generalAxis(x = qActual, maxVal = qMax, minVal = 0, 
                       tinyPlot = tinyPlot)
  xInfo <- generalAxis(x = localSeries$years, maxVal = decimal_date(end), 
                       minVal = decimal_date(start), padPercent = 0, tinyPlot = tinyPlot)
  
  line1 <- localINFO$shortName
  nameIstat <- c("minimum day", "7-day minimum", "30-day minimum", 
                 "median daily", "mean daily", "30-day maximum", "7-day maximum", 
                 "maximum day")
  
  line2 <-  paste0("\n", setSeasonLabelByUser(paStartInput = paStart, 
                                              paLongInput = paLong), "  ", nameIstat[istat])
  
  line3 <- paste0("\nSlope estimate is ",slopePct,"% per year, Mann-Kendall p-value is ",pValue)
  
  if(tinyPlot){
    title <- paste(nameIstat[istat])
  } else {
    title <- paste(line1, line2, line3)
  }
  
  if (!printTitle){
    title <- ""
  }
  
  genericEGRETDotPlot(x = localSeries$years, y = localSeries$qActual, 
                      xlim = c(xInfo$bottom, xInfo$top), ylim = c(yInfo$bottom, 
                                                                  yInfo$top), xlab = "", ylab = yLab, customPar = customPar, 
                      xTicks = xInfo$ticks, yTicks = yInfo$ticks, cex = cex, 
                      plotTitle = title, cex.axis = cex.axis, cex.main = cex.main, 
                      tinyPlot = tinyPlot, lwd = lwd, col = col, ...)
  lines(localSeries$years, localSeries$qSmooth, lwd = lwd, 
        col = col)
}

#########################################################################################
###### this the the function you will use to make the Quantile Kendall Plot #############
#########################################################################################

plotQuantileKendall <- function(eList, startDate = NA, endDate = NA, 
                                paStart = 4, paLong = 12,     
                                legendLocation = "topleft", legendSize = 1.0,
                                yMax = NA, yMin = NA) {
  localDaily <- eList$Daily
  localINFO <- eList$INFO
  localINFO$paStart <- paStart
  localINFO$paLong <- paLong
  start <- as.Date(startDate)
  end <- as.Date(endDate)
  
  if(is.na(startDate)){
    start <- as.Date(localDaily$Date[1]) 
  } 
  
  if(is.na(endDate)){
    end <- as.Date(localDaily$Date[length(localDaily$Date)])
  }
  
  localDaily <- subset(localDaily, Date >= start & Date <= end)
  eList <- as.egret(localINFO,localDaily)
  eList <- setPA(eList, paStart=paStart, paLong=paLong)
  
  v <- makeSortQ(eList)
  sortQ <- v[[1]]
  time <- v[[2]]
  results <- trendSortQ(sortQ, time)
  pvals <- c(0.001,0.01,0.05,0.1,0.25,0.5,0.75,0.9,0.95,0.99,0.999)
  zvals <- qnorm(pvals)
  name <- eList$INFO$shortName
  #  ymax <- trunc(max(results$slopePct)*10)
  #  ymax <- max(ymax + 2, 5)
  #  ymin <- floor(min(results$slopePct)*10)
  #  ymin <- min(ymin - 2, -5)
  #  yrange <- c(ymin/10, ymax/10)
  #  yticks <- axisTicks(yrange, log = FALSE)
  ymax <- max(results$slopePct + 0.5, yMax, na.rm = TRUE)
  ymin <- min(results$slopePct - 0.5, yMin, na.rm = TRUE)
  yrange <- c(ymin, ymax)
  yticks <- axisTicks(yrange, log = FALSE, nint =7)
  p <- results$pValueAdj
  color <- ifelse(p <= 0.1,"black","snow3")
  color <- ifelse(p < 0.05, "red", color)
  pvals <- c(0.001,0.01,0.05,0.1,0.25,0.5,0.75,0.9,0.95,0.99,0.999)
  zvals <- qnorm(pvals)
  name <- paste0("\n", eList$INFO$shortName,"\n",
                 start," through ", end, "\n", 
                 setSeasonLabelByUser(paStartInput = paStart, paLongInput = paLong))
  plot(results$z,results$slopePct,col = color, pch = 20, cex = 1.0, 
       xlab = "Daily non-exceedance probability", 
       ylab = "Trend slope in percent per year", 
       xlim = c(-3.2, 3.2), ylim = yrange, yaxs = "i", 
       las = 1, tck = 0.02, cex.lab = 1.2, cex.axis = 1.2, 
       axes = FALSE, frame.plot=TRUE)
  mtext(name, side =3, line = 0.2, cex = 1.2)
  axis(1,at=zvals,labels=pvals, las = 1, tck = 0.02)
  axis(2,at=yticks,labels = TRUE, las = 1, tck = 0.02)
  axis(3,at=zvals,labels=FALSE, las = 1, tck=0.02)
  axis(4,at=yticks,labels = FALSE, tick = TRUE, tck = 0.02)
  abline(h=0,col="blue")
  legend(legendLocation,c("> 0.1","0.05 - 0.1","< 0.05"),col = c("snow3",                                            "black","red"),pch = 20, title = "p-value",
         pt.cex=1.0, cex = legendSize * 1.5)
}    

#########################################################################################
############  This next function combines four individual trend graphs (for mimimum day,
########### median day, mean day, and maximum day) along with the quantile kendall graph
#########################################################################################

plotFiveTrendGraphs <- function(eList, startDate = NA, endDate = NA, 
                                paStart = 4, paLong = 12, qUnit = 2, window = 30, 
                                legendLocation = "topleft", legendSize = 1.0) {
  localDaily <- eList$Daily
  localINFO <- eList$INFO
  localINFO$paStart <- paStart
  localINFO$paLong <- paLong
  localINFO$window <- window
  
  start <- as.Date(startDate)
  end <- as.Date(endDate)
  
  if(is.na(startDate)){
    start <- as.Date(localDaily$Date[1]) 
  } 
  
  if(is.na(endDate)){
    end <- as.Date(localDaily$Date[length(localDaily$Date)])
  }
  
  localDaily <- subset(localDaily, Date >= start & Date <= end)
  
  eList <- as.egret(localINFO,localDaily)
  eList <- setPA(eList, paStart=paStart, paLong=paLong, window=window)
  # this next line of code is inserted so that when paLong = 12, we always use the
  # climate year when looking at the trends in the annual minimum flow
  paStart1 <- if(paLong == 12)  4 else paStart
  plotFlowTrend(eList, istat = 1, qUnit = qUnit, paStart = paStart1, paLong = paLong, window = window)
  plotFlowTrend(eList, istat = 4, qUnit = qUnit, paStart = paStart, paLong = paLong, window = window)
  plotFlowTrend(eList, istat = 8, qUnit = qUnit, paStart = paStart, paLong = paLong, window = window)
  plotFlowTrend(eList, istat = 5, qUnit = qUnit, paStart = paStart, paLong = paLong, window = window)
  # now the quantile kendall
  plotQuantileKendall(eList, startDate = startDate, endDate = endDate, paStart = paStart,
                      paLong = paLong, legendLocation = legendLocation, legendSize = legendSize)
  
} 

#########################################################################################
########### makeSortQ creates a matrix called Qsort. 
############It sorted from smallest to largest over dimDays 
############(if working with full year dimDays=365), 
#############and also creates other vectors that contain information about this array.
#########################################################################################

makeSortQ <- function(eList){
  localINFO <- getInfo(eList)
  localDaily <- getDaily(eList)
  paStart <- localINFO$paStart
  paLong <- localINFO$paLong
  # determine the maximum number of days to put in the array
  numDays <- length(localDaily$DecYear)
  monthSeqFirst <- localDaily$MonthSeq[1]
  monthSeqLast <- localDaily$MonthSeq[numDays]
  # creating a data frame (called startEndSeq) of the MonthSeq values that go into each year
  Starts <- seq(paStart, monthSeqLast, 12)
  Ends <- Starts + paLong - 1
  startEndSeq <- data.frame(Starts, Ends)
  # trim this list of Starts and Ends to fit the period of record
  startEndSeq <- subset(startEndSeq, Ends >= monthSeqFirst & Starts <= monthSeqLast)
  numYearsRaw <- length(startEndSeq$Ends)
  # set up some vectors to keep track of years
  good <- rep(0, numYearsRaw)
  numDays <- rep(0, numYearsRaw)
  midDecYear <- rep(0, numYearsRaw)
  Qraw <- matrix(nrow = 366, ncol = numYearsRaw)
  for(i in 1: numYearsRaw) {
    startSeq <- startEndSeq$Starts[i]
    endSeq <- startEndSeq$Ends[i]
    startJulian <- getFirstJulian(startSeq)
    # startJulian is the first julian day of the first month in the year being processed
    # endJulian is the first julian day of the month right after the last month in the year being processed
    endJulian <- getFirstJulian(endSeq + 1)
    fullDuration <- endJulian - startJulian
    yearDaily <- localDaily[localDaily$MonthSeq >= startSeq & (localDaily$MonthSeq <= endSeq), ]
    nDays <- length(yearDaily$Q)
    if(nDays == fullDuration) {
      good[i] <- 1
      numDays[i] <- nDays
      midDecYear[i] <- (yearDaily$DecYear[1] + yearDaily$DecYear[nDays]) / 2
      Qraw[1:nDays,i] <- yearDaily$Q
    }   else {
      numDays[i] <- NA
      midDecYear[i] <- NA
    }
  }
  # now we compress the matrix down to equal number of values in each column
  j <- 0
  numGoodYears <- sum(good)
  dayCounts <- ifelse(good==1, numDays, NA)
  lowDays <- min(dayCounts, na.rm = TRUE)
  highDays <- max(dayCounts, na.rm = TRUE)
  dimYears <- numGoodYears
  dimDays <- lowDays
  sortQ <- matrix(nrow = dimDays, ncol = dimYears)
  time <- rep(0,dimYears)
  for (i in 1:numYearsRaw){
    if(good[i]==1) {
      j <- j + 1
      numD <- numDays[i]
      x <- sort(Qraw[1:numD, i])
      # separate odd numbers from even numbers of days
      if(numD == lowDays) {
        sortQ[1:dimDays,j] <- x
      } else {
        sortQ[1:dimDays,j] <- if(odd(numD)) leapOdd(x) else leapEven(x)
      }
      time[j] <- midDecYear[i]
    } 
  }
  
  sortQList = list(sortQ,time)
  
  return(sortQList)         
}
#########################################################################################
########## Another function trendSortQ needed for Quantile Kendall
#########################################################################################

trendSortQ <- function(Qsort, time){
  # note requires packages zyp and rkt
  nFreq <- dim(Qsort)[1]
  nYears <- length(time)
  results <- as.data.frame(matrix(ncol=9,nrow=nFreq))
  colnames(results) <- c("slopeLog","slopePct","pValue","pValueAdj","tau","rho1","rho2","freq","z")
  for(iRank in 1:nFreq){
    mkOut <- rkt::rkt(time,log(Qsort[iRank,]))
    results$slopeLog[iRank] <- mkOut$B
    results$slopePct[iRank] <- 100 * (exp(mkOut$B) - 1)
    results$pValue[iRank] <- mkOut$sl
    outZYP <- zyp.zhang(log(Qsort[iRank,]),time)
    results$pValueAdj[iRank] <- outZYP[6]
    results$tau[iRank] <- mkOut$tau
    # I don't actually use this information in the current outputs, but the code is there 
    # if one wanted to look at the serial correlation structure of the flow series      
    serial <- acf(log(Qsort[iRank,]), lag.max = 2, plot = FALSE)
    results$rho1[iRank] <- serial$acf[2]
    results$rho2[iRank] <- serial$acf[3]
    frequency <- iRank / (nFreq + 1)
    results$freq[iRank] <- frequency
    results$z[iRank] <- qnorm(frequency)    
  }
  return(results)
}
#########################################################################################
################################## getFirstJulian finds the julian date of first day
################################## of a given month
#########################################################################################

getFirstJulian <- function(monthSeq){
  year <- 1850 + trunc((monthSeq - 1) / 12)
  month <- monthSeq - 12 * (trunc((monthSeq-1)/12))
  charMonth <- ifelse(month<10, paste0("0",as.character(month)), as.character(month))
  theDate <- paste0(year,"-",charMonth,"-01")
  Julian1 <- as.numeric(julian(as.Date(theDate),origin=as.Date("1850-01-01")))
  return(Julian1)
}

#########################################################################################
########### leapOdd  is a function for deleting one value 
############when the period that contains Februaries has a length that is an odd number
#########################################################################################

leapOdd <- function(x){
  n <- length(x)
  m <- n - 1
  mid <- (n + 1) / 2
  mid1 <- mid + 1
  midMinus <- mid - 1
  y <- rep(NA, m)
  y[1:midMinus] <- x[1:midMinus]
  y[mid:m] <- x[mid1:n]
  return(y)}

#########################################################################################
########### leapEven  is a function for deleting one value 
############when the period that contains Februaries has a length that is an even number
#########################################################################################

leapEven <- function(x){
  n <- length(x)
  m <- n - 1
  mid <- n / 2
  y <- rep(NA, m)
  mid1 <- mid + 1
  mid2 <- mid + 2
  midMinus <- mid - 1
  y[1:midMinus] <- x[1:midMinus]
  y[mid] <- (x[mid] + x[mid1]) / 2
  y[mid1:m] <- x[mid2 : n]
  return(y)
}

#########################################################################################
####### determines if the length of a vector is an odd number ###########################
#########################################################################################

odd <- function(x) {(!(x %% 2) == 0)}

#########################################################################################
########### calcWY calculates the water year and inserts it into a data frame
#########################################################################################


calcWY <- function (df) {
  df$WaterYear <- as.integer(df$DecYear)
  df$WaterYear[df$Month >= 10] <- df$WaterYear[df$Month >= 
                                                 10] + 1
  return(df)
}
#########################################################################################
##### calcCY calculates the climate year and inserts it into a data frame
#########################################################################################

calcCY <- function (df){
  df$ClimateYear <- as.integer(df$DecYear)
  df$ClimateYear[df$Month >= 4] <- df$ClimateYear[df$Month >= 
                                                    4] + 1
  return(df)
}
#########################################################################################
######## smoother is a function does the trend in real discharge units and not logs. 
######## It is placed here so that users wanting to run this alternative have it available
######## but it is not actually called by any function in this document
#########################################################################################

smoother <- function(xy, window){
  edgeAdjust <- TRUE
  x <- xy$x
  y <- xy$y
  n <- length(y)
  z <- rep(0,n)
  x1 <- x[1]
  xn <- x[n]
  for (i in 1:n) {
    xi <- x[i]
    distToEdge <- min((xi - x1), (xn - xi))
    close <- (distToEdge < window)
    thisWindow <- if (edgeAdjust & close) 
      (2 * window) - distToEdge
    else window
    w <- triCube(x - xi, thisWindow)
    mod <- lm(xy$y ~ x, weights = w)
    new <- data.frame(x = x[i])
    z[i] <- predict(mod, new)
  }
  return(z)
}

```


## Gwynns Falls at Villa Nova (GWVN)

Available daily Q: 1957-02-01 to present
Site code: 01589300 

```{r message=F,warning=F}

# Set gauging site number
siteNumber_GWVN <- '01589300' # Gwynns Falls at Villa Nova MD

# Set start & end dates for period
startDate_GWVN <- '1957-10-01'
endDate_GWVN <- '2022-09-30'

# Read in daily discharge (Q) data from USGS National Water Info System (NWIS)
Daily_GWVN <- readNWISDaily(siteNumber_GWVN, '00060', startDate_GWVN, endDate_GWVN)
# 00060 is the USGS parameter code for stream discharge

# Read in information about site & set short name - note ** interactive
INFO_GWVN <- readNWISInfo(siteNumber_GWVN, '00060') 
# Set short name to 'Gwynns Falls-Villa Nova'
# Set station abbreviation to 'GWVN'
# Set 'Stream flow, mean. daily' to 'DailyQ_cfs'
# Set abbreviation for discharge units (ft3/s) to 'cfs'

# Create eList -- EGRET variable
eList_GWVN <- as.egret(INFO_GWVN, Daily_GWVN, NA, NA)

plotFlowTrend(eList_GWVN, qUnit = 'cfs', paStart = 10, istat = 5) # see stat #s below; mean daily Q

###########QUANTILE KENDALL###############

Start <- 1957 # first year of Gwynns Falls - Villa Nova record
End <- 2019

# create start & end dates
Start.date <- paste(Start, '-10-01', sep='')
End.date <- paste(End, '-09-30', sep='')

plotQuantileKendall(eList_GWVN, startDate = Start.date, endDate = End.date, 
                    paStart = 10, paLong = 12,
                    legendLocation = "topright", legendSize = 0.5)
```

## Gunpowder Falls at Hoffmanville

Available daily Q: 2000-04-01 to 2022-10-18
Site code: 01581810  

```{r message=F,warning=F}
# Set gauging site number
siteNumber_GFH <- '01581810' # Gunpowder Falls at Hoffmanville

# Set start & end dates for period
startDate_GFH <- '2000-10-01'
endDate_GFH <- '2022-09-30'

# Read in daily discharge (Q) data from USGS National Water Info System (NWIS)
Daily_GFH <- readNWISDaily(siteNumber_GFH, '00060', startDate_GFH, endDate_GFH)
# 00060 is the USGS parameter code for stream discharge

# Read in information about site & set short name - note ** interactive
INFO_GFH <- readNWISInfo(siteNumber_GFH, '00060') 
# Set short name to 'Gunpowder Falls at Hoffmanville'
# Set station abbreviation to 'GFH'
# Set 'Stream flow, mean. daily' to 'DailyQ_cfs'
# Set abbreviation for discharge units (ft3/s) to 'cfs'

# Create eList -- EGRET variable
eList_GFH <- as.egret(INFO_GFH, Daily_GFH, NA, NA)

plotFlowTrend(eList_GFH, qUnit = 'cfs', paStart = 10, istat = 5) # see stat #s below; mean daily Q

###########QUANTILE KENDALL###############

Start <- 2000 # first year of record
End <- 2019

# create start & end dates
Start.date <- paste(Start, '-10-01', sep='')
End.date <- paste(End, '-09-30', sep='')

plotQuantileKendall(eList_GFH, startDate = Start.date, endDate = End.date, 
                    paStart = 10, paLong = 12,
                    legendLocation = "topright", legendSize = 0.5)

```

## Piney Run at Dover (PRD)

Available daily Q: 1982-05-10 to present 
Site code: 01583100 

```{r message=F,warning=F}
# Set gauging site number
siteNumber_PRD <- '01583100' # Piney Run at Dover

# Set start & end dates for period
startDate_PRD <- '1982-10-01'
endDate_PRD <- '2022-09-30'

# Read in daily discharge (Q) data from USGS National Water Info System (NWIS)
Daily_PRD<- readNWISDaily(siteNumber_PRD, '00060', startDate_PRD, endDate_PRD)
# 00060 is the USGS parameter code for stream discharge

# Read in information about site & set short name - note ** interactive
INFO_PRD <- readNWISInfo(siteNumber_PRD, '00060') 
# Set short name to 'Piney Run-Dover'
# Set station abbreviation to 'PRD'
# Set 'Stream flow, mean. daily' to 'DailyQ_cfs'
# Set abbreviation for discharge units (ft3/s) to 'cfs'

# Create eList -- EGRET variable
eList_PRD <- as.egret(INFO_PRD, Daily_PRD, NA, NA)

plotFlowTrend(eList_PRD, qUnit = 'cfs', paStart = 10, istat = 4) # see stat #s below; mean daily Q

###########QUANTILE KENDALL###############

Start <- 1982 # first year of record
End <- 2019

# create start & end dates
Start.date <- paste(Start, '-10-01', sep='')
End.date <- paste(End, '-09-30', sep='')

plotQuantileKendall(eList_PRD, startDate = Start.date, endDate = End.date, 
                    paStart = 10, paLong = 12,
                    legendLocation = "topright", legendSize = 0.5)
```

## Morgan Run (MR)

Available daily Q: 1982-10-01 to present
Site code: 01586610 

```{r message=F,warning=F}
# Set gauging site number
siteNumber_MR <- '01586610' # Morgan Run

# Set start & end dates for period
startDate_MR <- '1982-10-01'
endDate_MR <- '2022-09-30'

# Read in daily discharge (Q) data from USGS National Water Info System (NWIS)
Daily_MR<- readNWISDaily(siteNumber_MR, '00060', startDate_MR, endDate_MR)
# 00060 is the USGS parameter code for stream discharge

# Read in information about site & set short name - note ** interactive
INFO_MR <- readNWISInfo(siteNumber_MR, '00060') 
# Set short name to 'Morgan Run-Louisville'
# Set station abbreviation to 'MR'
# Set 'Stream flow, mean. daily' to 'DailyQ_cfs'
# Set abbreviation for discharge units (ft3/s) to 'cfs'

# Create eList -- EGRET variable
eList_MR <- as.egret(INFO_MR, Daily_MR, NA, NA)

plotFlowTrend(eList_MR, qUnit = 'cfs', paStart = 10, istat = 4) # see stat #s below; mean daily Q

###########QUANTILE KENDALL###############

Start <- 1982 # first year of record
End <- 2019

# create start & end dates
Start.date <- paste(Start, '-10-01', sep='')
End.date <- paste(End, '-09-30', sep='')

plotQuantileKendall(eList_MR, startDate = Start.date, endDate = End.date, 
                    paStart = 10, paLong = 12,
                    legendLocation = "topright", legendSize = 0.5)
```

## Western Run 

Available daily Q: 1944-10-01 to present
Site code: 01583500   

```{r message=F,warning=F}
# Set gauging site number
siteNumber_WR <- '01583500' # Western Run 

# Set start & end dates for period
startDate_WR <- '1944-10-01'
endDate_WR <- '2022-09-30'

# Read in daily discharge (Q) data from USGS National Water Info System (NWIS)
Daily_WR<- readNWISDaily(siteNumber_WR, '00060', startDate_WR, endDate_WR)
# 00060 is the USGS parameter code for stream discharge

# Read in information about site & set short name - note ** interactive
INFO_WR <- readNWISInfo(siteNumber_WR, '00060') 
# Set short name to 'Western Run'
# Set station abbreviation to 'WR'
# Set 'Stream flow, mean. daily' to 'DailyQ_cfs'
# Set abbreviation for discharge units (ft3/s) to 'cfs'

# Create eList -- EGRET variable
eList_WR <- as.egret(INFO_WR, Daily_WR, NA, NA)

plotFlowTrend(eList_WR, qUnit = 'cfs', paStart = 10, istat = 5) # see stat #s below; mean daily Q

###########QUANTILE KENDALL###############

Start <- 1944 # first year of record
End <- 2019

# create start & end dates
Start.date <- paste(Start, '-10-01', sep='')
End.date <- paste(End, '-09-30', sep='')

plotQuantileKendall(eList_WR, startDate = Start.date, endDate = End.date, 
                    paStart = 10, paLong = 12,
                    legendLocation = "topright", legendSize = 0.5)

```


## LITTLE FALLS AT BLUE MOUNT, MD

Available daily Q: 1944-06-01 to present
Site code: 01582000    

```{r message=F,warning=F}
# Set gauging site number
siteNumber_LF <- '01582000' # Little Falls

# Set start & end dates for period
startDate_LF <- '1944-10-01'
endDate_LF <- '2022-09-30'

# Read in daily discharge (Q) data from USGS National Water Info System (NWIS)
Daily_LF<- readNWISDaily(siteNumber_LF, '00060', startDate_LF, endDate_LF)
# 00060 is the USGS parameter code for stream discharge

# Read in information about site & set short name - note ** interactive
INFO_LF <- readNWISInfo(siteNumber_LF, '00060') 
# Set short name to 'Little Falls at Blue Mount'
# Set station abbreviation to 'LF'
# Set 'Stream flow, mean. daily' to 'DailyQ_cfs'
# Set abbreviation for discharge units (ft3/s) to 'cfs'

# Create eList -- EGRET variable
eList_LF <- as.egret(INFO_LF, Daily_LF, NA, NA)

plotFlowTrend(eList_LF, qUnit = 'cfs', paStart = 10, istat = 5) # see stat #s below; mean daily Q

###########QUANTILE KENDALL###############

Start <- 1944 # first year of record
End <- 2019

# create start & end dates
Start.date <- paste(Start, '-10-01', sep='')
End.date <- paste(End, '-09-30', sep='')

plotQuantileKendall(eList_LF, startDate = Start.date, endDate = End.date, 
                    paStart = 10, paLong = 12,
                    legendLocation = "topright", legendSize = 0.5)


```