---
title: "Part 2 Ambient WC"
author: "Frank"
date: "`r format(Sys.Date())`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.width=18, fig.height=12)
```

```{css, echo=FALSE}
h1, h4 {
  text-align: center;
  color: black;
  font-weight: bold;
}

/* Whole document: */
body{
  font-family: Times New Roman;
  font-size: 16pt;
}
/* Headers */
h1,h2,h3,h4,h5,h6{
  font-size: 24pt;
}

<style type="text/css">
body, td {
   font-size: 16px;
}
code.r{
  font-size: 16px;
}
pre {
  font-size: 16px
}
</style>
  
```

# Introduction

These are codes to analyze and explore patterns in part 2 ambient water chemistry data. 

## Import data

```{r message=F,warning=F}
library(readxl)
partb_wc <- read_excel("F:/One Drive - Towson U/Thesis stuff/Data/Water chemistry data/Part b WC data/Ambient WC data.xlsx", 
    sheet = "Combined")
```

# Range of data

First, let's summarize the data. 

```{r message=F,warning=F}
library(dplyr)
partb_wc %>% summary()
```

# Investigate the relationship between pH and %Do

The relationship between %Do and pH is expected to be positive. As more photosynthesis happens in stream, more CO2 is pumped out, which then leads to higher %Do and higher pH. Let's investigate to see if the relationship holds. 

```{r message=F,warning=F}
library(dplyr)
ph_do_clean <- as.data.frame(partb_wc) %>% 
  rename(c(Temp=`Temp (C)`, SC = `SC (mS/cm)`, TSS = `TSS (ppm)`, 
          DO_ppm = `DO (ppm)`, DO_p =  `DO (%)`, Sal =  `Salinity (ppt)`)) %>% 
  filter(!is.na(DO_p)) # (1) rename columns; (2) filter NA data from DO column; 

cor.test(ph_do_clean$pH,ph_do_clean$DO_p, method = "pearson") #correlational test between pH and DO using Pearson cor
cor.test(ph_do_clean$pH,ph_do_clean$DO_p, method = "spearman") #correlational test between pH and DO using Spearman cor

library(ggplot2)
ph_do_clean %>% ggplot(aes(pH, DO_p)) + geom_point()
```

In conclusion, pH and %DO have no correlation with each other in this data set, since there seems to be some extreme values. 

# Create a matrix of scatterplots

## Pearson

These codes were used to create a scatterplot matrix (Pearson's correlation) using the package GGally (Schloerke et al. 2021). 

```{r message=F, warning=F}
library(dplyr)
library(GGally)

lowerFn <- function(data, mapping, method = "lm", ...) {
  p <- ggplot(data = data, mapping = mapping) +
    geom_point(colour = "blue") +
    geom_smooth(method = method, color = "red", ...)
  p
} #code adopted from WCC - an user on Stack Overflow. Check out his profile here: https://stackoverflow.com/users/1102552/wcc.

wc_clean <- partb_wc %>% rename(c(Temp=`Temp (C)`, SC = `SC (mS/cm)`, TSS = `TSS (ppm)`, 
          DO_ppm = `DO (ppm)`, DO_p =  `DO (%)`, Sal =  `Salinity (ppt)`)) %>% 
  filter(!is.na(SC) & !is.na(TSS) & !is.na(DO_p)) 

wc_clean %>% 
  ggpairs(columns = c("Temp","SC","TSS","DO_ppm","DO_p","pH","Sal"), 
          lower = list(continuous = wrap(lowerFn, method = "lm")),
  diag = list(continuous = wrap("barDiag", colour = "blue")),
  upper = list(continuous = wrap("cor", size = 10))
) +  theme(axis.text.y = element_text(colour = "black", size = 12, face = "bold"), 
    axis.text.x = element_text(colour = "black", face = "bold", size = 12), 
    legend.text = element_text(size = 12, face ="bold", colour ="black"), 
    axis.title.x = element_text(face = "bold", size = 14, colour = "black"), 
    legend.title = element_text(size = 14, colour = "black", face = "bold"), 
    panel.background = element_blank(), panel.border = element_rect(colour = "black", fill = NA, size = 1.2),
    legend.key=element_blank(),
    plot.title = element_text(color = "black", size = 30, face = "bold", hjust = 0.5)) + 
   labs(
    title = "Correlation matrix of ambient WC (Pearson)")
  
```

## Spearman

Previous codes will be modified to display Spearman correlation. For now, traditional method will be applied. Package *corrr* (Kuhn et al. 2020) will be used to test for significant correlations among variables.

```{r message=F,warning=F}
library(corrr)

wc_cor <- wc_clean[,3:ncol(wc_clean)] %>% correlate(use = "pairwise.complete.obs", method = "spearman") %>% 
  select(Temp, SC, TSS, DO_ppm, DO_p, pH, Sal) %>% mutate(Variable = c(1,2,3,4,5,6,7), .before = Temp)
wc_cor$Variable <- recode(wc_cor$Variable, 
                          "1" = "Temp", 
                          "2" = "SC",
                          "3" = "TSS",
                          "4" = "DO_ppm",
                          "5" = "DO_p",
                          "6" = "pH",
                          "7" = "Sal")
wc_cor
```

Another way of doing this, but will also give you a correlation and p-value matrix. Package *rstatix* will be used (Kassambara 2021.)

```{r message=F,warning=F}
library(rstatix)
options(digits = 3)   # report 3 significant digits
wc_clean[,3:ncol(wc_clean)] %>% cor_mat(Temp, SC, TSS, DO_ppm, DO_p, pH, Sal, method = "spearman") %>% cor_get_pval()
```
## Temp vs pH

It seems like the relationship between water temperature and pH is a little odd. In the previous code chunk, NA data were removed, thus, led to some temperature and pH data being removed as well. Now I am testing correlation between temp vs pH alone.

```{r message=F,warning=F}
library(dplyr)
temp_ph <- as.data.frame(partb_wc) %>% 
  rename(c(Temp=`Temp (C)`, SC = `SC (mS/cm)`, TSS = `TSS (ppm)`, 
          DO_ppm = `DO (ppm)`, DO_p =  `DO (%)`, Sal =  `Salinity (ppt)`)) %>% 
  filter(!is.na(Temp) & !is.na(pH)) # (1) rename columns; (2) filter NA data from temp and pH column;

cor.test(temp_ph$pH,temp_ph$Temp, method = "pearson") #correlational test between pH and DO using Pearson cor
cor.test(temp_ph$pH,temp_ph$Temp, method = "spearman") #correlational test between pH and DO using Spearman cor

```

Results showed that the correlation between temperature and pH is not statisitcally significant with a wide range of 95% CI. 


# References
1. Barret Schloerke, Di Cook, Joseph Larmarange, Francois Briatte, Moritz Marbach, Edwin Thoen, Amos Elberg and Jason Crowley
  (2021). GGally: Extension to 'ggplot2'. R package version 2.1.2. https://CRAN.R-project.org/package=GGally
2. Max Kuhn, Simon Jackson and Jorge Cimentada (2020). corrr: Correlations in R. R package version 0.4.3.
  https://CRAN.R-project.org/package=corrr
3.  Alboukadel Kassambara (2021). rstatix: Pipe-Friendly Framework for Basic Statistical Tests. R package version 0.7.0.
  https://CRAN.R-project.org/package=rstatix

